\documentclass[11pt]{article}
\usepackage{fullpage}

\begin{document}
\pagestyle{empty}
\parindent=0pt

\section*{\centering Problem Set 1}

\section{Cyclotron}

Suppose you have a (non-relativistic) electron spinning circles in a magnetic field.

\subsection{Show that the frequency of oscillation does not depend on the velocity of the electron.}
\vspace{0.25in}
\subsection{In what waveband would an electron spiralling in the Earth's magnetic field emit?}
\vspace{0.25in}

\section{Brightnesses, Magnitudes, and Photons}

In the old days, astronomers classified the brightness of celestial
objects using the magnitude system, which roughly corresponded to the
sensitivity of the human eye. Today we continue to use this ancient system
because\dots well, I don't know, but we do. The magnitude of
an object in a given filter $i$ is
\begin{equation}m_i = −2.5 \log_{10}(F_i/F_{0,i}) + m_{0,i}
\end{equation}
where $F_i$ is the observed flux integrated over a given filter\begin{equation}
F_i = \int_0^\infty{F_\nu(\nu)\phi(\nu)~d\nu}
\end{equation}
where $\phi(\nu)$ is the filter transmission function (a number
between 0 and 1).The magnitude system is calibrated by specifying $F_{0,i}$ and $m_{0,i}$ which
may be done in a number of ways. A common system isVega-magnitudes, in which the star Vega is defined to have $m\approx0$ in all bands.

\subsection{}
Grab the filter transmission functions and the spectrum of Vega(bessel\_V.dat and vega\_spectrum.dat in the github repository, noting that
the Vega spectrum is given as $F_\lambda$, with units ergs s$^{-1}$ cm$^{-2}$ \AA$^{-1}$).
numerically integrate the apparent V-band magnitude of Vega to determine how many
photons the world's largest ground-based optical telescope (Keck, with a mirror diameter of 10m)
would collect per second from Vega.

\subsection{}
Use the fact that Vega is about 8 pc away, and has a diameter of approximately 2.5 $R_\odot$, to
determine the specific intensity of Vega in cgs units, assuming Vega appears as a uniform disk on the sky.

\subsection{}
If we were twice as close to Vega (say, 4 pc), figure how many photons per second Keck would receive
in V band, and re-determine the specific intensity of Vega.  Are the scalings of photons per second and
specific intensity with distance consistent with one another?  How so?

\section{Dust Bowl}

Dust particles are about 100 microns in diameter.  In the years of the Dust
Bowl in the 1930s, it was not uncommon for dust clouds to completely obscure
the Sun.  According to Wikipedia:
\begin{quote}
On April 14, 1935, known as ``Black Sunday", 20 of the worst ``black
blizzards" occurred across the entire sweep of the Great Plains, from Canada
south to Texas. The dust storms caused extensive damage and turned the day to
night; witnesses reported they could not see five feet in front of them at
certain points.
\end{quote}

\subsection{}
Using this observation, estimate the number density of dust particles in the air.
We'll do a more sophisticated
treatment of dust grains later in the class, but for now, assume these dust grains to be perfectly
opaque, and large enough to be treated geometrically (without diffraction, etc.).

\subsection{}
If these clouds extended half a mile into the air, how much topsoil was lost to each one of these clouds?

\end{document}
