\documentclass[11pt]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{titlesec}
\titleformat{\subsection}[runin]{\normalfont\large\bfseries}{\thesubsection}{1em}{}

\def\Te{{T_e}}
\def\Tp{{T_p}}
\def\tep{{t_{ep}}}
\def\trec{{t_{rec}}}
\begin{document}
\pagestyle{empty}
%\parindent=0pt

\section*{\centering Problem Set 7}

\section{Pulsar Dispersion Measure}

A radio astronomer uses the Green Bank Telescope to observe a pulsar
in a band from 1 to 2 GHz, recording a series of frequency power spectra for 1 s. The resulting
(noisy) frequency spectrum is recorded as pulsar.dat (available on the course GitHub repo).  It looks like
noise, but
I promise you that, for the right dispersion measure, there is a pulse in there.

\subsection{}
De-disperse the measured power spectra to determine the
DM of the pulsar.  You may have to try many DMs to find the one that works.
To help you, the beginning of the file is timed to the arrival time without dispersion delay.  For bonus
points, you could make your code not depend on this fact.

Just a warning: be careful about the difference between $\omega$ and $\nu$.  The reported frequencies in
the data file are $\nu$, of course.

\subsection{}
For an assumed density of electrons in the interstellar medium of 0.03 cm$^{-3}$, calculate the
distance to this pulsar. Does this seem reasonable (what scale would you use to judge ``reasonable")?

\subsection{}
For such an assumed electron density, is the radio astronomer safely observing above the plasma cut-off frequency?

\subsection{}
Calculate the optical depth to Thomson scattering along the line-of-sight to this pulsar.

\section{Faraday Rotation}

Consider the propagation of light thorugh a magnetized plasma.  The magnetic field is uniform: $\vec B_0=B_0\hat z$.
Light travels parallel to $\hat z$.

An electron in the plasma feels a force from the electromagnetic wave and a force from the externally imposed
$B$-field.  Its equation of motion reads
\begin{equation}
m\dot \vec v=-e\vec E-\frac{e}{c}\vec v\times \vec B_0
\end{equation}
where the electric field $\vec E$ can be decomp[osed into right-circularly polarized (RCP) and
left-circularly-polarized (LCP) waves:
\begin{equation}
\vec E=E_0(\hat x\mp i\hat y)e^{i(k_\mp z-\omega t)}
\end{equation}
where it is understood that the real part of $\vec E$ should be taken.  The upper sign ($-$) corresponds
to RCP waves, while the lower sign ($+$) corresponds to LCP waves.

In the equation of motion above, we have neglected the Lorentz force from the wave's
$B$-field, since it is small (by $v/c$) compared to the force from the wave's $E$-field.

\subsection{}
Prove that the solution of the equation of motion reads
\begin{equation}
\vec v=\frac{-ie}{m(\omega\pm\omega_{cyc})}\vec E
\end{equation}
where $\omega_{cyc}\equiv eB_0/mc$.  This is the hardest part of the derivation for the dispersion
relation of RCP and LCP waves.

\subsection{}
Based on Rybicki \& Lightman Problem 8.3.

The signal from a pulsed, polarized source is measured to have an arrival time delay that varies
with frequency as $dt_p/d\omega=1.1\cdot10^{-5}~{\rm s}^2$, and a Faraday rotation that
varies with frequency as $d\Delta\theta/d\omega=1.9\cdot10^{-4}~{\rm s}$.  The measurements are
made around the frequency $\omega=10^8~{\rm s}^{-1}$, and the source is at unknown distance from
the earth.  Find the mean magnetic field, $\langle B_\parallel\rangle$, in the interstellar space
between the earth and the source:
\begin{equation}
\langle B_\parallel\rangle\equiv\frac{\int{n B_\parallel~ds}}{\int{n~ds}}
\end{equation}

\end{document}
